---
title: "Semi quantitative proteomics enables mapping of global neutrophil dynamics following influenza virus infection"
date: "2019-01-25"
author: Chuanxin Liu
categories:
- Research
- Coding
tags:
- research
- publication
image: "https://8heddw.ch.files.1drv.com/y4mSLrpvgdAbg0tTkx3RhzfRqpLs7L6xDT3U5NO7USxOLxnX0XjJFEg5j90FI-Xw9qovOyAX-e92EKz2hUT8QWsdLX2vB5J0KsxqxeWhgZQvT2O0LKq9YezyECWyCl5mZcAF_x2V5YpMXPUsK0VmA4wvUZui9_bQxt8RlEtQx-XqixRUR6kcIWCULI5HnhGhgQs?width=1920&height=1080&cropmode=none"
output:
  blogdown::html_page:
    dev: "svg" ## set the default output device to svg
    toc: true
---

## Source code and dataset availability

Source code is hosted on [github](https://github.com/codetrainee/Neutrophil-systems-biology)

## Prerequisite

We first loaded libraries required for analysis. Scripts used in the analysis could be found in `data/Rscripts`. Packages and their version are specificed in a plain file as well as listed at the bottom of the page. 

```{r source-R-scripts, warning=FALSE}
source("data/Rscripts/load_libraries.R")
source("data/Rscripts/utilities.R")
```

The theme for figures generated by ggplot2 was set to `dust` style from [ggthemr](https://github.com/cttobin/ggthemr) packages.

```{r set-the-global-theme-to-ggthemr-style}
ggthemr("dust")
```

To have a tidy output, the figures were centered and message and warning were suppressed by defaults. Code was also hidden for simplicity.

```{r global-options-for-knitr}
knitr::opts_chunk$set(fig.align = "center",
                      message   = FALSE,
                      warning   = FALSE,
                      cache     = TRUE,
                      echo      = FALSE,
                      results   = "hide")
```

All data were stored under the below folders and can be found in the [Source code and dataset availability](##Source-code-and-dataset-availability)

```{r set-global-variables}
path <- "data/Neutrophil-proteomics-data-analysis-following-influenza-or-bacterial-infection/"
```

In raw files generated from [MaxQuant](https://www.maxquant.org/) software, a plenty of `NA` values has to be removed before downstream analyses. To survive the filtering step, a pick-up protein should meet following requirements:

- should be detected in at least one comparison group.
- should be picked up in all of three independent replicates.

## Analysis of neutrophil proteome at homeostasis or following a lethal IAV infection

3 experiments were included in the analysis pipeline. The first two were combined together to identify neutrophil-derived markers in a 1000 pfu PR8 influenza infection compared to PBS treated control. The third experiment compared the neutrophil proteome in different doses of influenza infection and bacterial infection.

### Raw data cleaning

We first loaded datasets from two experiments and removed the potential contaminations marked by MaxQuant.

```{r load-raw-data-for-first-exp, warning=FALSE, results='hide'}
raw_data_nov <-
  glue(path, "20170428-Proteomics-201611-Requant.txt") %>% clean_rawdata()

raw_data_dec <-
  glue(path, "20170428-Proteomics-201612-Requant.txt") %>% clean_rawdata()
```

The datasets were annotated as below. Missing values were filtered out using customised functions.

| Labelling  | Sample                        |
|------------|-------------------------------|
| Light (L)  | 1000 PFU PR8 BM neutrophil    |
| Medium (M) | 1000 PFU PR8 Blood neutrophil |
| Heavy (H)  | 1000 PFU PR8 BAL neutrophil   |

Table: **Experiment 1 Biological replicate 1**

| Labelling  | Sample                        |
|------------|-------------------------------|
| Light (L)  | 1000 PFU PR8 BAL neutrophil   |
| Medium (M) | 1000 PFU PR8 BM neutrophil    |
| Heavy (H)  | 1000 PFU PR8 Blood neutrophil |

Table: **Experiment 1 Biological replicate 2**

| Labelling  | Sample                        |
|------------|-------------------------------|
| Light (L)  | 1000 PFU PR8 Blood neutrophil |
| Medium (M) | 1000 PFU PR8 BAL neutrophil   |
| Heavy (H)  | 1000 PFU PR8 BM neutrophil    |

Table: **Experiment 1 Biological replicate 3**

| Labelling  | Sample                     |
|------------|----------------------------|
| Light (L)  | PBS BM neutrophil          |
| Medium (M) | PBS Blood neutrophil       |
| Heavy (H)  | 1000 PFU PR8 BM neutrophil |

Table: **Experiment 2 Biological replicate 1**

| Labelling  | Sample                     |
|------------|----------------------------|
| Light (L)  | 1000 PFU PR8 BM neutrophil |
| Medium (M) | PBS BM neutrophil          |
| Heavy (H)  | PBS Blood neutrophil       |

Table: **Experiment 2 Biological replicate 2**

| Labelling  | Sample                  |
|------------|-------------------------|
| Light (L)  | PBS Blood neutrophil    |
| Medium (M) | 1000 PFU PR8 neutrophil |
| Heavy (H)  | PBS BM neutrophil       |

Table: **Experiment 2 Biological replicate 3**

```{r data-precleaning, warings=FALSE}
valid_list <- c("tissue", "1", "2", "3")

data_nov <-
  raw_data_nov %>%
  ## select the required columns
  dplyr::select(matches("Gene names|normalized Rep")) %>%
  magrittr::set_colnames(c(
    "Gene_names",
    "C_1000_PFU_PR8_Blood_vs_1000_PFU_PR8_BM_1",
    "B_1000_PFU_PR8_BAL_vs_1000_PFU_PR8_BM_2_invert",
    "A_1000_PFU_PR8_BAL_vs_1000_PFU_PR8_Blood_3",
    "B_1000_PFU_PR8_BAL_vs_1000_PFU_PR8_BM_1",
    "A_1000_PFU_PR8_BAL_vs_1000_PFU_PR8_Blood_2_invert",
    "C_1000_PFU_PR8_Blood_vs_1000_PFU_PR8_BM_3_invert",
    "A_1000_PFU_PR8_BAL_vs_1000_PFU_PR8_Blood_1",
    "C_1000_PFU_PR8_Blood_vs_1000_PFU_PR8_BM_2",
    "B_1000_PFU_PR8_BAL_vs_1000_PFU_PR8_BM_3_invert"
  )) %>%
  clean_data() %>% 
  validate_df(valid_list)

data_nov_stat <- remove_invalid(data_nov)

## Annotations were similar to the preparation step for the above experiment.
data_dec <-
  raw_data_dec %>%
  dplyr::select(matches("Gene names|normalized Rep")) %>%
  magrittr::set_colnames(c(
    "Gene_names",
    "c_PBS_Blood_vs_PBS_BM_1",
    "b_1000_PFU_PR8_BM_vs_PBS_BM_2_invert",
    "a_1000_PFU_PR8_BM_vs_PBS_Blood_3",
    "b_1000_PFU_PR8_BM_vs_PBS_BM_1",
    "a_1000_PFU_PR8_BM_vs_PBS_Blood_2_invert",
    "c_PBS_Blood_vs_PBS_BM_3_invert",
    "a_1000_PFU_PR8_BM_vs_PBS_Blood_1",
    "c_PBS_Blood_vs_PBS_BM_2",
    "b_1000_PFU_PR8_BM_vs_PBS_BM_3_invert"
  )) %>%
  clean_data() %>% 
  validate_df(valid_list)

data_dec_stat <- remove_invalid(data_dec)
```

Proteins passed the filtering step were subjected to statistics tests for difference analysis. A protein with a fold change greater than 2 and p value less than 0.05 in at least one comparison group was considered as a significant one in protein abundance change.
With all data cleaned and processed, visualization was the next step to clearly illustrate the proteomics profiles.

```{r combine-two-exp-and-apply-statistics-test}
## combine two experiments
all_stat <- dplyr::inner_join(data_nov_stat, data_dec_stat) 

## Indirect way to compute the ratio in terms of comparison between 1000 pfu PR8 Blood and PBS Blood.
all_stat <- 
  map(c("1", "2", "3"),
      function(x) {
        colname <- sym(glue("X_1000_PFU_PR8_Blood_vs_PBS_Blood_{x}"))
        grep_string <- glue("C_1000_PFU_PR8_Blood_vs_1000_PFU_PR8_BM_{x}|a_1000_PFU_PR8_BM_vs_PBS_Blood_{x}")
        subset <- dplyr::select(all_stat, matches(grep_string, ignore.case = FALSE))

        mutate(all_stat, 
               !!colname := apply(subset,
                              1,
                              prod))
      }) %>% 
  purrr::reduce(left_join) %>% 
  mutate_at(vars(-matches("Gene_names")), list(~log2))
  
sig_list <- c("A", "B", "C", "b", "c", "X")

## calculate the p value and group mean
all_stat <-
  map(sig_list,
      function(x) {
        p_value <- sym(glue("{x}_p_value"))
        mean_value <- sym(glue("{x}_mean"))
        grep_string <- glue("^{x}.+[0-9]$")
        subset <- dplyr::select(all_stat, matches(grep_string, ignore.case = FALSE))
        
        mutate(
          all_stat,
          !!mean_value := apply(subset, 1, mean),
          !!p_value := apply(subset, 1, function(x)
            t.test(x)$p.value)
        )
      }) %>%
  purrr::reduce(left_join) 

## filter the protein that pass the statistic test
all_stat <-
  map(sig_list,
      function(x) {
        ratio_sig <- sym(glue("{x}_sig"))
        p_value <- sym(glue("{x}_p_value"))
        mean_value <- sym(glue("{x}_mean"))
        
        mutate(all_stat, !!ratio_sig := !!p_value < 0.05 &
                 abs(!!mean_value) > 1)
      }) %>%
  purrr::reduce(left_join) %>% 
  ## remove the comparison between 1000 pfu PR8 BM versus PBS Blood, which makes no sense in the following analysis
  dplyr::select(-matches("1000_PFU_PR8_BM_vs_PBS_Blood", ignore.case = FALSE))

## Comparison between 1000 pfu PR8 Blood and PBS Blood was included.
all_stat_with_flu_blood_vs_pbs_blood <-
  all_stat %>%
  mutate(Sig_diff = rowSums(dplyr::select(., matches("_sig$", ignore.case = FALSE))))

## Exclude the comparsion between 1000 pfu PR8 Blood and PBS Blood.
all_stat <-
  all_stat %>%
  dplyr::select(-matches("^X_")) %>%
  mutate(Sig_diff = rowSums(dplyr::select(., matches("_sig$"))))
```

### Visualization of pick-up proteins in independent experiments and biological replicates

Before moving on, we checked all proteins identified in all replicates and visualized them via Venn Diagram. As shown, a total of 1372 proteins survived in the filtering step and they were used for subsequent analyses.

```{r Venn-Diagram, fig.dim=c(8, 8)}
## Visualise the pick-up difference between independent replicates and experiments.
## for the first experiment
data_nov_replicate <-
  data_nov %>%
  filter_at(vars(matches("rep.*valid")), any_vars(. == 1)) %>%
  dplyr::select(matches("^rep_[0-9]{1}_valid$")) %>%
  ## Venn diagram was drawn with [eulerr](https://github.com/jolars/eulerr)
  euler(.) %>%
  plot(.,
       main = "Experiment 1",
       quantities = TRUE,
       fill = c("#f1a340", "#f7f7f7", "#998ec3"),
       legend = list(side = "bottom")
  )

## for the second experiment
data_dec_replicate <-
  data_dec %>%
  filter_at(vars(matches("rep.*valid")), any_vars(. == 1)) %>%
  dplyr::select(matches("^rep_[0-9]{1}_valid$")) %>%
  euler(.) %>%
  plot(.,
       main = "Experiment 2",
       quantities = TRUE,
       fill = c("#d8b365", "#f5f5f5", "#5ab4ac"),
       legend = list(side = "bottom")
  )


## combine two experiment
all_replicate <-
  full_join(
    data_nov %>%
      mutate(Nov = apply(dplyr::select(., matches("rep.*valid")), 1, function(x) all(x == 1))) %>%
      dplyr::filter(Nov == TRUE) %>%
      dplyr::select(Gene_names, Nov),
    data_dec %>%
      mutate(Dec = apply(dplyr::select(., matches("rep.*valid")), 1, function(x) all(x == 1))) %>%
      dplyr::filter(Dec == TRUE) %>%
      dplyr::select(Gene_names, Dec)
  ) %>%
  dplyr::select(-Gene_names)
all_replicate[is.na(all_replicate)] <- FALSE

all_replicate <-
  all_replicate %>%
  euler(.) %>%
  plot(.,
       main = "Two experiments combined",
       quantities = TRUE,
       fill = c("#fc8d59", "#91bfdb"),
       legend = list(side = "bottom")
  )

## customise the layout of venn diagrams using [customLayout](https://github.com/zzawadz/customLayout)
Venn_lay <- lay_new(
  matrix(1:2, nc = 2),
  widths = c(1, 1)
)

Venn_lay2 <- lay_new(matrix(1))

Venn_cl <- lay_bind_row(Venn_lay, Venn_lay2, heights = c(1.2, 1))

lay_grid(list(data_nov_replicate, data_dec_replicate, all_replicate), Venn_cl)
```

### Overall impression of proteome dataset

Data reproducibility between biological replicates were examined by a PCA analysis. The comparison between 1000 pfu PR8 Blood vs PBS Blood was included. From the PCA plot, we can clearly see that the comparison between 1000 pfu PR8 blood and PBS blood was similar to that of 1000 pfu PR8 BM versus PBS BM. Another similar comparison was also identified in 1000 pfu PR8 blood vs 1000 pfu PR8 BM and PBS blood and PBS BM.

```{r PCA-analysis-with-flu-blood-vs-pbs-blood, fig.keep="all", fig.dim=c(8, 8)}
set.seed(123)
all_stat_pca <-
  all_stat_with_flu_blood_vs_pbs_blood %>%
  dplyr::select(matches("_vs_")) %>%
  as.matrix() %>%
  prcomp(.)

## PCA results QC
fviz_eig(all_stat_pca)

## caculate the variances explained by each principle component
all_stat_pca_pov <- all_stat_pca$sdev^2 / sum(all_stat_pca$sdev^2)

## extract the dataframe from PCA analysis
all_stat_pca_df <- all_stat_pca$rotation %>%
  as.data.frame() %>%
  mutate(Sample = row.names(.),
         Comparison = str_replace_all(str_extract(Sample, "(?<=[A-Za-z]{1}_).+(?=_[0-9]{1})"), "_", " "),          batch = str_extract(Sample, "[0-9]{1}$"))

## visualise the PCA using ggplot2
ggplot(all_stat_pca_df, aes(PC1, PC2, colour = Comparison)) +
  geom_point(size = 4) +
  labs(
    x = sprintf("PC1: %.2f %% Variance", all_stat_pca_pov[[1]] * 100),
    y = sprintf("PC2: %.2f %% Variance", all_stat_pca_pov[[2]] * 100)
  ) +
  scale_colour_manual(values = brewer.pal(6, "RdBu")[c(1:3, 6, 4, 5)]) +
  theme_classic() +
  theme(aspect.ratio = 1,
        legend.position = "bottom",
        panel.border = element_rect(fill = "transparent"),
        legend.direction = "vertical"
        )
```

As the comparison between 1000 pfu PR8 blood vs PBS blood was calculated *in silico*, it was removed temporarily from the following analysis unless indicated. We redrawn the PCA plot with the aforementioned group removed.

```{r PCA-analysis, fig.keep="all", fig.dim=c(8, 8)}
set.seed(123)
all_stat_pca <-
  all_stat %>%
  dplyr::select(matches("_vs_")) %>%
  as.matrix() %>%
  prcomp(.)

## PCA results QC
fviz_eig(all_stat_pca)

## caculate the variances explained by each principle component
all_stat_pca_pov <- all_stat_pca$sdev^2 / sum(all_stat_pca$sdev^2)

## extract the dataframe from PCA analysis
all_stat_pca_df <- all_stat_pca$rotation %>%
  as.data.frame() %>%
  mutate(Sample = row.names(.),
         Comparison = str_replace_all(str_extract(Sample, "(?<=[A-Za-z]{1}_).+(?=_[0-9]{1})"), "_", " "),          batch = str_extract(Sample, "[0-9]{1}$"))

## visualise the PCA using ggplot2
ggplot(all_stat_pca_df, aes(PC1, PC2, colour = Comparison)) +
  geom_point(size = 4) +
  labs(
    x = sprintf("PC1: %.2f %% Variance", all_stat_pca_pov[[1]] * 100),
    y = sprintf("PC2: %.2f %% Variance", all_stat_pca_pov[[2]] * 100)
  ) +
  scale_colour_manual(values = brewer.pal(6, "RdBu")) +
  theme_classic() +
  theme(aspect.ratio = 1,
        legend.position = "bottom",
        panel.border = element_rect(fill = "transparent"),
        legend.direction = "vertical"
        )
```

The heatmap for all proteins was also generatef for an overall impression of expression profiles. The first one included the *in silico* calculated group.

```{r heatmap-for-all-proteins-with-flu-blood-vs-pbs-blood, fig.dim=c(6, 8)}
## heatmaps were generated by [ComplexHeatmap](https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html)
## construct the matrix for heatmap input
all_stat_heatmap_matrix <-
  all_stat_with_flu_blood_vs_pbs_blood %>%
  dplyr::select(matches("Gene|_vs_")) %>%
  as.matrix() %>%
  `rownames<-`(.[, "Gene_names"]) %>%
  `colnames<-`(sapply(colnames(.), function(x) str_replace(x, "^[A-Za-z]{1}_", ""))) %>%
  `colnames<-`(sapply(colnames(.), function(x) str_replace_all(x, "_", " "))) %>%
  .[, !colnames(.) %in% c("Gene names")] %>%
  `class<-`("numeric")

## Same with heatmap for sig proteins
all_stat_heatmap_ha_df <-
  data.frame(type = colnames(all_stat_heatmap_matrix)) %>%
  mutate(Comparison = str_extract(type, ".+(?= [0-9]{1}$)")) %>%
  dplyr::select(-type)

## construct the customised annotation
all_stat_heatmap_ha <-
  columnAnnotation(
    df = all_stat_heatmap_ha_df,
    col = list(Comparison = setNames(
      brewer.pal(6, "RdBu")[c(1:3, 6, 4, 5)],
      sort(unique(all_stat_heatmap_ha_df$Comparison))
    ) %>% sort(.)),
    annotation_legend_param = list(Comparison = list(
      nrow = 3,
      title = "Comparison",
      title_gp = gpar(fontsize = 8),
      labels_gp = gpar(fontsize = 6)
    ))
  )

## set the seed for reproducibility
set.seed(123)

all_stat_heatmap <-
  Heatmap(all_stat_heatmap_matrix,
    top_annotation = all_stat_heatmap_ha,
    heatmap_legend_param = list(
      title = "Log2 Fold Change",
      title_gp = gpar(fontsize = 8),
      labels_gp = gpar(fontsize = 6),
      legend_direction = "horizontal",
      legend_width = unit(1.5, "in")
    ),
    show_column_names = FALSE,
    show_row_names = FALSE,
    show_row_dend = FALSE
  )

draw(all_stat_heatmap, heatmap_legend_side = "bottom", annotation_legend_side = "bottom")

for (an in colnames(all_stat_heatmap_ha_df)) {
  decorate_annotation(an, {
    ## annotation names on the left
    grid.text(an,
      unit(1, "npc") + unit(2, "mm"),
      0.6,
      default.units = "npc",
      just = "left",
      gp = gpar(fontsize = 8)
    )
  })
}
```

The next heatmap was without the *in silico* calculated group.

```{r heatmap-for-all-proteins, fig.dim=c(6, 8)}
## heatmaps were generated by [ComplexHeatmap](https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html)
## construct the matrix for heatmap input
all_stat_heatmap_matrix <-
  all_stat %>%
  dplyr::select(matches("Gene|_vs_")) %>%
  as.matrix() %>%
  `rownames<-`(.[, "Gene_names"]) %>%
  `colnames<-`(sapply(colnames(.), function(x) str_replace(x, "^[A-Za-z]{1}_", ""))) %>%
  `colnames<-`(sapply(colnames(.), function(x) str_replace_all(x, "_", " "))) %>%
  .[, !colnames(.) %in% c("Gene names")] %>%
  `class<-`("numeric")

## Same with heatmap for sig proteins
all_stat_heatmap_ha_df <-
  data.frame(type = colnames(all_stat_heatmap_matrix)) %>%
  mutate(Comparison = str_extract(type, ".+(?= [0-9]{1}$)")) %>%
  dplyr::select(-type)

## construct the customised annotation
all_stat_heatmap_ha <-
  columnAnnotation(
    df = all_stat_heatmap_ha_df,
    col = list(Comparison = setNames(
      brewer.pal(6, "RdBu")[1:5],
      sort(unique(all_stat_heatmap_ha_df$Comparison))
    ) %>% sort(.)),
    annotation_legend_param = list(Comparison = list(
      nrow = 3,
      title = "Comparison",
      title_gp = gpar(fontsize = 8),
      labels_gp = gpar(fontsize = 6)
    ))
  )

## set the seed for reproducibility
set.seed(123)

all_stat_heatmap <-
  Heatmap(all_stat_heatmap_matrix,
    top_annotation = all_stat_heatmap_ha,
    heatmap_legend_param = list(
      title = "Log2 Fold Change",
      title_gp = gpar(fontsize = 8),
      labels_gp = gpar(fontsize = 6),
      legend_direction = "horizontal",
      legend_width = unit(1.5, "in")
    ),
    show_column_names = FALSE,
    show_row_names = FALSE,
    show_row_dend = FALSE
  )

draw(all_stat_heatmap, heatmap_legend_side = "bottom", annotation_legend_side = "bottom")

for (an in colnames(all_stat_heatmap_ha_df)) {
  decorate_annotation(an, {
    ## annotation names on the left
    grid.text(an,
      unit(1, "npc") + unit(2, "mm"),
      0.6,
      default.units = "npc",
      just = "left",
      gp = gpar(fontsize = 8)
    )
  })
}
```

We then submitted the identified proteins to DAVID database to investigate the cellular compartments they belong to.

```{r david-GO-analysis}
## cell compartment GO analysis
all_stat_gene <- all_stat %>% pull(Gene_names)
all_stat_gene_ENTREZID <- mapIds(org.Mm.eg.db,
                                 keys = all_stat_gene,
                                 column = "ENTREZID",
                                 keytype = "SYMBOL",
                                 multiVals = "first") %>% na.omit

## gene list was searched against DAVID database https://david.ncifcrf.gov
david_gene <- readr::read_tsv(glue(path, "all-stat-gene-DAVID-database-GO-cell-compartment.txt"))

## save it locally
## write_tsv(david_gene, glue(path, "all-stat-gene-DAVID-database-GO-cell-compartment.txt"))

p_go_cell_compartment <-
  ggplot(david_gene %>%
         filter(FDR < 0.05) %>%
         arrange(Count) %>%
         mutate(Term = factor(Term, Term)),
       aes(Term, Count)) +
  geom_bar(stat = "identity",
           fill = "#cccccc") +
  scale_y_continuous(breaks = seq(0, 1500, 300),
                     limits = c(0, 1600),
                     expand = c(0, 0)) +
  coord_flip()

p_go_cell_compartment
```

### Analysis of proteins with significant abundance changes

Proteins with significant expression changes across at least one comparison group were filterd out. The 155 protein expression matrix was subjected to `fviz_nbclust` using `wss` method to determine the optimal cluster for visualization.

```{r potential-cluster-number, fig.dim=c(6, 6)}
## Annotation is same with heatmap for all proteins
sig_stat_heatmap_df <-
  all_stat %>%
  filter(Sig_diff > 0) %>%
  dplyr::select(matches("Gene|_vs_")) %>%
  `colnames<-`(sapply(colnames(.), function(x) str_replace(x, "^[A-Za-z]{1}_", ""))) %>%
  `colnames<-`(sapply(colnames(.), function(x) str_replace_all(x, "_", " ")))

## determine the potential cluster numbers
fviz_nbclust(sig_stat_heatmap_df %>% dplyr::select(-matches("Gene")),
  kmeans,
  method = "wss"
)
```

Based on the figure, 9 clusters were the best one to separate protein expression difference.

```{r, heatmap-for-signifant, fig.dim=c(6, 15)}
set.seed(123)
sig_stat_heatmap_df <-
  sig_stat_heatmap_df %>%
  mutate(cluster = kmeans(sig_stat_heatmap_df %>% dplyr::select(-matches("Gene")), 9)$cluster)

sig_stat_heatmap_matrix <-
  sig_stat_heatmap_df %>%
  as.matrix() %>%
  `rownames<-`(.[, "Gene names"] %>% toupper) %>%
  .[, !colnames(.) %in% c("Gene names")] %>%
  `class<-`("numeric")


sig_stat_heatmap_ha_row <-
  rowAnnotation(
    df = sig_stat_heatmap_df %>% dplyr::select(cluster) %>% as.data.frame(),
    col = list(cluster = setNames(
      brewer.pal(9, "BrBG"),
      unique(sig_stat_heatmap_df$cluster) %>% sort(.)
    )),
    width = unit(0.2, "in"),
    annotation_legend_param = list(
      nrow = 1,
      title = "Cluster",
      title_gp = gpar(fontsize = 8),
      labels_gp = gpar(fontsize = 8),
      legend_width = unit(1, "in")
    )
  )

sig_stat_heatmap <-
  Heatmap(sig_stat_heatmap_matrix[, !colnames(sig_stat_heatmap_matrix) %in% c("cluster")],
          top_annotation = all_stat_heatmap_ha,
          heatmap_legend_param = list(
            title = "Log2 Fold Change",
            title_gp = gpar(fontsize = 8),
            labels_gp = gpar(fontsize = 6),
            legend_direction = "horizontal",
            legend_width = unit(1.5, "inch")
          ),
          show_column_names = FALSE,
          show_row_dend = FALSE,
          show_row_names = TRUE,
          row_names_gp = gpar(fontsize = 6),
          combined_name_fun = NULL,
          split = sig_stat_heatmap_matrix[, "cluster"],
          gap = unit(1.5, "mm")
  )

draw(sig_stat_heatmap_ha_row + sig_stat_heatmap, heatmap_legend_side = "bottom", annotation_legend_side = "bottom")

for (an in colnames(all_stat_heatmap_ha_df)) {
  decorate_annotation(an, {
    ## annotation names on the left
    grid.text(an,
              unit(1, "npc") + unit(2, "mm"),
              0.6,
              default.units = "npc",
              just = "left",
              gp = gpar(fontsize = 8)
    )
  })
}
```

All analyses above were only about all proteins from different comparison group. To identify proteins in each comparison group, we visualized data with volcano plots. The dataset was first cleaned for the subsequent visualization and the volcano plots were shown below.

```{r volcano-plot-data-cleaning}
Comparison_list <- c("A", "B", "C", "b", "c", "X")
all_stat_volcano <- 
  map(Comparison_list, 
      function(x) {
        p_value <- sym(glue("{x}_p_value"))
        mean_value <- sym(glue("{x}_mean"))
        colname <- sym(x)
        
        mutate(all_stat_with_flu_blood_vs_pbs_blood,
               !!colname := case_when(
                 !!p_value < 0.05 & !!mean_value > 1 ~ "up",
                 !!p_value < 0.05 & !!mean_value < -1 ~ "down",
                 TRUE ~ "ns"
               ))
      }
      ) %>% 
  purrr::reduce(left_join)
```

```{r volcano-plot-function, cache=FALSE}
## customized function for generating a series of volcano plots.
geom_volcano <- function(data = data, type = type, gene = gene_list)
{
  x = sym(paste0(type, "_mean"))
  y = sym(paste0(type, "_p_value"))
  type = sym(type)

  title = str_extract(colnames(data), sprintf("^%s_.+", toString(type))) %>% unique() %>% str_extract(., "(?<=[A-Za-z]_).+(?=_[0-9]$)") %>% str_replace_all(., "_", " ") %>% na.omit()
  data <- mutate_at(data, vars(!!y), list(~-log10(.)))
  x_limits <- c(-6, 6)
  y_limits <- c(0, 6)
  x_breaks = seq(-6, 6, 2)
  y_breaks = seq(0, 6, 1)
  point_colour = setNames(c(brewer.pal(11, "RdBu")[2], "gray50", brewer.pal(11, "RdBu")[10]),
                          c("up", "ns", "down"))
  down_regulated_number = data %>% filter(!!x <= -1, !!y > -log10(0.05)) %>%  nrow(.)
  up_regulated_number = data %>% filter(!!x >= 1, !!y > -log10(0.05)) %>%  nrow(.)

  ggplot(data, aes(!!x, !!y)) +
    geom_point(aes(color = !!type), size = 2) +
    geom_vline(xintercept = c(-1, 1),
               linetype = "dashed",
               color = "gray50",
               size = 0.25) +
    geom_hline(yintercept = -log10(0.05),
               linetype =  "dashed",
               color = "gray50",
               size = 0.25) +
    geom_text_repel(data =  data %>% filter(Gene_names %in% gene, !!type == "down" | !!type == "up"),
                    aes(label = Gene_names),
                    nudge_y      = 7,
                    direction    = "x",
                    angle        = 90,
                    vjust        = 1,
                    segment.size = 0.2
                    ) +
    annotate("text",
             x = -Inf,
             y = 0.1,
             label = paste("Downregulated gene: ", down_regulated_number),
             hjust = 0,
             size = 4) +
    annotate("text",
             x = Inf,
             y = 0.1,
             label = paste("Upregulated gene: ", up_regulated_number),
             hjust = 1,
             size = 4) +
    labs(title = title, x = "Log2 Fold Change", y = "-Log10(p value)") +
    scale_x_continuous(limits = x_limits, breaks = x_breaks) +
    scale_y_continuous(limits = y_limits, breaks = y_breaks) +
    scale_color_manual(values = point_colour, guide = FALSE) +
    theme_classic() +
    theme(
          axis.line.x = element_line(size = 0.25),
          axis.line.y = element_line(size = 0.25),
          axis.ticks.x  = element_line(size = 0.25),
          axis.ticks.y  = element_line(size = 0.25),
          axis.title = element_text(size = 10),
          axis.text = element_text(size = 8),
          plot.title = element_text(size = 12,
                               face = "plain"))
}
```

```{r volcano-plot-list, fig.dim=c(20, 15), cache=FALSE}
gene_list <- c(
  "Oas3",
  "Isg15",
  "Ifi204",
  "Ifit1",
  "Gbp2",
  "Gbp7",
  "Cmpk2",
  "Ddx58",
  "Gvin1",
  "Selenbp1;Selenbp2",
  "Rnf213",
  "Lgals9",
  "Dtx3l",
  "Fth1",
  "Il1rn",
  "Mmp9",
  "Mmp8",
  "Lrg1",
  "F5",
  "Retnlg",
  "Ltf",
  "Il1f9;Il36g",
  "Lcn2",
  "Mcm2",
  "Dnmt1",
  "Lig1",
  "Btf3",
  "Abt1",
  "Actbl2",
  "haemaglobin alpha 2;Hba",
  "Thbs1",
  "Tubb1",
  "Tpm4",
  "Phospho1",
  "Srgn",
  "Stfa2l1;Stfa2",
  "Asprv1",
  "Ftl1;Ftl2",
  "H2-K1",
  "Slfn1",
  "Por",
  "Plek",
  "Pcna",
  "Cebpe",
  "Ckap4",
  "Asns",
  "Top2a",
  "Rrm1",
  "Mcmbp",
  "Crispld2",
  "Sorl1",
  "Timp2",
  "Olfm4",
  "Ngp",
  "Pglyrp1",
  "Tcn2",
  "Alb",
  "Chi3l1",
  "Hp",
  "Itih5",
  "Qsox1"
)

volcano_g <- lapply(c("A", "B", "C", "b", "c", "X"), function(x) geom_volcano(data = all_stat_volcano,
                                                                         type = x,
                                                                         gene = gene_list))

plot_grid(plotlist = volcano_g[c(1, 2, 4, 3, 5, 6)], nrow = 2)
```

We are also interested in the pathways these proteins involved. Therefore, we filtered the proteins and annotated them as down or up. They were used to search the Reactome database respectively and retrieved the pathway data for visualization.

```{r pathway-analysis}
## as of 16 Oct 2018
all_stat_pathway <-
  all_stat %>%
  as_tibble() %>%
  dplyr::filter(Sig_diff == TRUE & !(abs(c_mean) >= 1 & c_p_value < 0.05)) %>%
  dplyr::select(Gene_names, A_mean, A_p_value, C_mean, C_p_value, b_mean, b_p_value) %>%
  mutate(`1000_pfu_PR8_BAL_vs_1000_pfu_PR8_Blood_direction` = ifelse(A_mean < -1 & A_p_value < 0.05, "down", ifelse(A_mean > 1 & A_p_value < 0.05, "up", "ns")),
         `1000_pfu_PR8_Blood_vs_1000_pfu_PR8_BM_direction` = ifelse(C_mean < -1 & C_p_value < 0.05, "down", ifelse(C_mean > 1 & C_p_value < 0.05, "up", "ns")),
         `1000_pfu_PR8_BM_vs_pbs_BM_direction` = ifelse(b_mean < -1 & b_p_value < 0.05, "down", ifelse(b_mean > 1 & b_p_value < 0.05, "up", "ns")))

purrr::map(c("1000_pfu_PR8_BAL_vs_1000_pfu_PR8_Blood", "1000_pfu_PR8_Blood_vs_1000_pfu_PR8_BM", "1000_pfu_PR8_BM_vs_pbs_BM"),
           function(x) {
             direction <- sym(glue::glue(x, "_direction"))

             assign(glue::glue(x, "_down"),
                    all_stat_pathway %>%
                      dplyr::filter(!!direction == "down") %>%
                      pull(Gene_names),
                    envir = .GlobalEnv)

             assign(glue::glue(x, "_up"),
                    all_stat_pathway %>%
                      dplyr::filter(!!direction == "up") %>%
                      dplyr::pull(Gene_names),
                    envir = .GlobalEnv)
           }) %>% rm(.)
```

Because in blood tissues, many proteins related to ribisome were downregulated and they would interfere the final outputs. To address this, we separated them first and proceeded to visualization.

```{r pathway-data-cleaning}
pathway_file <- glue::glue(path, "/20181018-reactome-pathway-analysis.xlsx")

Comparison_name_conversion <-
  setNames(c("1000_pfu_PR8_BAL_vs_1000_pfu_PR8_Blood",
             "1000_pfu_PR8_Blood_vs_1000_pfu_PR8_BM",
             "1000_pfu_PR8_BM_vs_pbs_BM"),
           c("flu_bal_vs_flu_blood",
             "flu_blood_vs_flu_bm",
             "flu_bm_vs_pbs_bm"))

Pathway_all <-
  purrr::map((readxl::excel_sheets(pathway_file)),
    function(i){
      assign(glue::glue("pathway_", i),
             readxl::read_excel(pathway_file, sheet = i) %>%
               dplyr::select(`Pathway name`, `#Entities found`, `Entities FDR`, `Submitted entities found`) %>%
               magrittr::set_colnames(c("Pathway", "Found", "FDR", "Submitted_found")) %>%
               dplyr::filter(FDR < 0.05 & Found > 2) %>%
               mutate(Entries = str_split(Submitted_found, ";"),
                      Down = map(1:nrow(.), ~ get(glue::glue(Comparison_name_conversion[i], "_down"))),
                      Up = map(1:nrow(.), ~ get(glue::glue(Comparison_name_conversion[i], "_up")))) %>%
               mutate(Down_no = map2(.$Entries, .$Down, ~ intersect(.x, .y)) %>% map(length) %>% as.numeric(),
                      Up_no = map2(.$Entries, .$Up, ~ intersect(.x, .y)) %>% map(length) %>% as.numeric(),
                      Database_extra_no = Found - Down_no - Up_no,
                      Comparison = rep(i, nrow(.))),
             envir = .GlobalEnv)
      }
    ) %>%
  purrr::reduce(bind_rows)

## There are lots of ribosome-related proteins in the blood compartments.
## To ease visualization, we separated the
## pathways into two groups.
## - pathways involves the ribosome proteins
## - pathways excludes the ribosome proteins
Pathway_all_Blood <-
  Pathway_all %>%
  dplyr::filter(grepl("^Rp", Submitted_found)) %>%
  ##  dplyr::filter(!grepl("^Rp", Submitted_found)) %>%
  dplyr::filter(!Pathway %in% c("Immune System")) %>%
  mutate(p_y = setNames(1:length(.$Pathway %>% unique), .$Pathway %>% unique %>% sort(decreasing = T))[Pathway] %>% as.numeric(),
         c_x = setNames(1:length(.$Comparison %>% unique), .$Comparison %>% unique %>% sort(decreasing = T))[Comparison] %>% as.numeric()) %>%
  mutate(radius = sqrt(Found/40))

Pathway_all_exclude_Blood <-
  Pathway_all %>%
  dplyr::filter(!grepl("^Rp", Submitted_found)) %>%
  dplyr::filter(!Pathway %in% c("Immune System")) %>%
  mutate(p_y = setNames(1:length(.$Pathway %>% unique), .$Pathway %>% unique %>% sort(decreasing = T))[Pathway] %>% as.numeric(),
         c_x = setNames(1:length(.$Comparison %>% unique), .$Comparison %>% unique %>% sort(decreasing = T))[Comparison] %>% as.numeric()) %>%
  mutate(radius = sqrt(Found/40))
```

We first checked the proteins without ribisome "contamination".

```{r pathway-exclude-ribosomes, fig.dim = c(10, 15)}
g_pathway_exclue_Blood <-
  ggplot(data = Pathway_all_exclude_Blood) +
  geom_scatterpie(aes(x = c_x, y = p_y, r = radius), color = "transparent", cols = c("Down_no", "Up_no", "Database_extra_no"), data = Pathway_all_exclude_Blood) +
  scale_y_continuous(limits = c(-2, 36), breaks = seq(1, 35, 1), labels = unique(Pathway_all_exclude_Blood$Pathway) %>% sort(decreasing = T))  +
  scale_x_continuous(labels = c("1000 pfu PR8 BM vs pbs BM",
                                "1000 pfu PR8 Blood vs 1000 pfu PR8 BM",
                                "1000 pfu PR8 BAL vs 1000 pfu PR8 Blood"
                              ),
                   breaks = c(1, 2, 3)) +
  coord_equal()  +
  scale_fill_manual(values = c("#cccccc", "#2c7bb6", "#d7191c")) +
  geom_scatterpie_legend(Pathway_all_exclude_Blood$radius,
                         x = 1,
                         y = -2,
                         labeller=function(x) x=c(3, 7, 11, 15)
                         ) +
  theme(axis.text = element_text(size = 8),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab(NULL) +
  xlab(NULL)

g_pathway_exclue_Blood
```

And then the proteins related to ribisomes were visualized.

```{r pathway-include-ribosomes, fig.dim = c(10, 15)}
g_pathway_Blood <-
  ggplot(data = Pathway_all_Blood) +
  geom_scatterpie(aes(x = c_x + 1, y = p_y, r = radius), color = "transparent", cols = c("Down_no", "Up_no", "Database_extra_no"), data = Pathway_all_Blood) +
  scale_y_continuous(limits = c(-2, 40), breaks = seq(1, 39, 1), labels = unique(Pathway_all_Blood$Pathway) %>% sort(decreasing = T))  +
  scale_x_continuous(labels = c("",
                                "1000 pfu PR8 Blood vs 1000 pfu PR8 BM",
                                ""
                                ),
                     limits = c(0.5, 3.5),
                     breaks = c(1, 2, 3)) +
  coord_equal()  +
  scale_fill_manual(values = c("#cccccc", "#2c7bb6", "#d7191c")) +
  geom_scatterpie_legend(Pathway_all_Blood$radius,
                         x = 1,
                         y = -2,
                         labeller=function(x) x=c(3, 7, 11, 15)) +
  theme(axis.text = element_text(size = 8),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab(NULL) +
  xlab(NULL)

g_pathway_Blood
```

Finally we used the alluvival flow to show the protein abundance changes from PBS BM to 1000 pfu PR8 BM and then Blood and BAL.

```{r ggalluvial, fig.dim=c(6, 15)}
g_alluvial <-
  ggplot(all_stat_pathway %>%
         dplyr::select(1, 8:10) %>%
         gather(key, value, 2:4) %>%
         mutate(freq = 1,
                value = factor(value, levels = c("up", "ns", "down")),
                key = factor(key, levels = c("1000_pfu_PR8_BM_vs_pbs_BM_direction",
                                             "1000_pfu_PR8_Blood_vs_1000_pfu_PR8_BM_direction",
                                             "1000_pfu_PR8_BAL_vs_1000_pfu_PR8_Blood_direction"))),
       aes(key, freq, stratum = value, alluvium = Gene_names, fill = value, label = Gene_names)) +
  geom_stratum(alpha = .5) +
  geom_flow() +
  geom_text(stat = "alluvium", size = 3) +
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1),
        plot.margin=margin(5,5,5,50,unit="pt")) +
  labs(x = NULL, y = NULL) +
  scale_x_discrete(labels = c("1000 pfu PR8 BM vs PBS BM",
                              "1000 pfu PR8 Blood vs 1000 pfu PR8 BM",
                              "1000 pfu PR8 BAL vs 1000 pfu PR8 Blood"))

g_alluvial
```

### iBAQ intensity ranking

The intensity data were extracted from iBAQ values and only the proteins identified in both experiments were used. Ranking all proteins based on their iBAQ intensity values is a good alternative to visualize their abundance in a cell when no abosulte copy number is available.

```{r intensity-data-preparation}
## customise a function to convert `0` to `NA` that is compatible with our previous defined function.
is_zero <- function(x) { ifelse(x == 0, TRUE, FALSE)}

data_nov_intensity <-
  raw_data_nov %>%
  dplyr::select(matches("Gene names|iBAQ\\s[A-Z]{1}\\sRep[0-9]+$")) %>%
  magrittr::set_colnames(c("Gene_names",
                           "A_1st_1000_PFU_PR8_BM_1",
                           "B_1st_1000_PFU_PR8_Blood_1",
                           "C_1st_1000_PFU_PR8_BAL_1",
                           "C_1st_1000_PFU_PR8_BAL_2",
                           "A_1st_1000_PFU_PR8_BM_2",
                           "B_1st_1000_PFU_PR8_Blood_2",
                           "B_1st_1000_PFU_PR8_Blood_3",
                           "C_1st_1000_PFU_PR8_BAL_3",
                           "A_1st_1000_PFU_PR8_BM_3")) %>%
  clean_data()
  

## To make the iBAQ dataset compatible with our `filter_valid` function.
data_nov_intensity[is_zero(data_nov_intensity)] <- NA

valid_list <- c("tissue", "1", "2", "3")

data_nov_intensity <-
  data_nov_intensity %>%
  validate_df(valid_list) %>% 
  remove_invalid() %>% 
  tidyr::gather(group, value, -Gene_names)

data_dec_intensity <-
  raw_data_dec %>%
  dplyr::select(matches("Gene names|iBAQ\\s[A-Z]{1}\\sRep[0-9]+$")) %>%
  magrittr::set_colnames(c(
    "Gene_names",
    "F_2nd_PBS_BM_1",
    "E_2nd_PBS_Blood_1",
    "D_2nd_1000_PFU_PR8_BM_1",
    "D_2nd_1000_PFU_PR8_BM_2",
    "F_2nd_PBS_BM_2",
    "E_2nd_PBS_Blood_2",
    "E_2nd_PBS_Blood_3",
    "D_2nd_1000_PFU_PR8_BM_3",
    "F_2nd_PBS_BM_3"
  )) %>%
  clean_data() 

## To make the iBAQ dataset compatible with our `filter_valid` function.
data_dec_intensity[is_zero(data_dec_intensity)] <- NA

data_dec_intensity <-
  data_dec_intensity %>%
  validate_df(valid_list) %>% 
  remove_invalid() %>% 
  tidyr::gather(group, value, -Gene_names)

data_intensity <-
  bind_rows(data_nov_intensity, data_dec_intensity) %>%
  mutate(gene_count = 1) %>%
  group_by(Gene_names) %>%
  mutate(gene = sum(gene_count)) %>%
  dplyr::filter(gene == 18) %>%
  dplyr::select(-gene_count, -gene) %>%
  group_by(group) %>%
  mutate(ranking = dplyr::dense_rank(-value),
         percent = value / sum(value)) %>%
  arrange(-percent, group) %>%
  mutate(percent_cum = cumsum(percent))
```

```{r customised-function-for-intensity-ranking-visualization}
## Customise a function to draw ranking plots and insert a small plot which contains proteins accounting for 75% of total abundance.
geom_ranking <- function(df) {
  ranking_marker <- df %>% dplyr::slice(base::which.min(abs(percent_cum - 0.75))) %>% pull(ranking)

  df_annotation <- df %>% dplyr::filter(ranking <= ranking_marker)

  title <- df$group %>% unique() %>% toString() %>% stringr::str_replace_all("_", " ") %>% stringr::str_replace_all("^[A-F]\\s", "")

  p <-
    ggplot(df, aes(ranking, percent_cum)) +
    geom_line() +
    geom_point() +
    labs(title = title,
         y = NULL,
         x = NULL) +
    scale_x_continuous(breaks = seq(0, 2000, 200),
                       labels = seq(0, 2000, 200),
                       limits = c(0, 2000)) +
    scale_y_continuous(breaks = seq(0, 1, 0.1)) +
    theme(axis.title = element_text(size = 10),
          axis.text = element_text(size = 8),
          plot.title = element_text(size = 12,
                                    face = "plain",
                                    hjust = 0.5))

  x_max <- df_annotation %>% pull(ranking) %>% max()
  y_max <- df_annotation %>% dplyr::filter(ranking == x_max) %>% pull(percent_cum)

  p_annotation <-
    ggplotGrob(
      ggplot(df_annotation, aes(ranking, percent_cum)) +
        geom_line() +
        geom_point() +
        geom_segment(aes(x = 0,
                         xend = x_max,
                         y = y_max,
                         yend = y_max),
                     linetype = "dashed") +
        geom_segment(aes(x = x_max,
                         xend = x_max,
                         y = 0,
                         yend = y_max,),
                     linetype = "dashed") +
        scale_x_continuous(limits = c(0, 80),
                           breaks = c(seq(0, 40, 10), x_max, 80),
                           labels = c(seq(0, 40, 10), x_max, 80),
                         expand = expand_scale(mult = c(0, 0.01),
                                               add = c(0, 0))) +
        scale_y_continuous(breaks = seq(0, 0.80, 0.1),
                           expand = expand_scale(mult = c(0, 0.05),
                                         add = c(0, 0))) +
        expand_limits(c(0, 0)) +
        labs(y = NULL,
             x = NULL) +
        theme_bw() +
        theme(axis.title = element_text(size = 10),
              axis.text = element_text(size = 8))
    )

  p + annotation_custom(
    grob = p_annotation,
    xmin = 500,
    xmax = 1900,
    ymin = 0.1,
    ymax = 0.85
  )
}
```

```{r visualised-intensity-ranking, fig.dim=c(16, 20)}
ranking_plot <-
  purrr::map(
    data_intensity$group %>% unique() %>% stringr::str_sort(),
    ~ geom_ranking(dplyr::filter(data_intensity, group == .x))
)

cowplot::plot_grid(plotlist = ranking_plot,
                   ncol = 3,
                   scale = 0.9) + ## reduce this for a bit more space
  cowplot::draw_label("Ranking",
                      x = 0.5,
                      y = 0,
                      vjust = -0.5,
                      angle = 0) +
  cowplot::draw_label("Cumulative proportion",
                      x = 0,
                      y = 0.5,
                      vjust = 1.5,
                      angle = 90)
```

## Analysis of BAL neutrophil proteome changes following different doses of influenza infection or bacterial infection

### Raw data cleaning

The above data analysis revealed that neutrophils underwent proteomics profile changes during influenza infection. To investigate whether it is a pathogen specific or a dose-dependent response during infection. We performed another round of proteomics experiment where the design was shown below:

| Labelling  | Sample                                                |
| :--------- | :-----------------------------------------------------|
| Light (L)  | 100 pfu PR8 BAL neutrophil                            |
| Medium (M) | 1000 pfu PR8 BAL neutrophil                           |
| Heavy (H)  | 3 × 10^6^ CFU *Pseudomonas Aeruginosa* BAL neutrophil |

Table: **Experiment 3 Biological replicate 1**

| Labelling  | Sample                                                |
| :--------- | :---------------------------------------------------- |
| Light (L)  | 1000 PFU PR8 BAL neutrophil                           |
| Medium (M) | 3 × 10^6^ CFU *Pseudomonas Aeruginosa* BAL neutrophil |
| Heavy (H)  | 100 pfu PR8 BAL neutrophil                            |

Table: **Experiment 3 Biological replicate 2**

| Labelling  | Sample                                                |
| :--------- | :-----------------------------------------------------|
| Light (L)  | 3 × 10^6^ CFU *Pseudomonas Aeruginosa* BAL neutrophil |
| Medium (M) | 100 PFU PR8 BAL neutrophil                            |
| Heavy (H)  | 1000 PFU PR8 BAL neutrophil                           |

Table: **Experiment 3 Biological replicate 3**

We first tried data cleaning for this dataset. The proteins with different abundance were also filtered out.

```{r raw-data-cleaning-for-the-third-exp, warning=FALSE}
data_201712 <-
  glue("{path}/20180206-Proteomics-201801-Requant.txt") %>% 
  clean_rawdata() %>% 
  dplyr::select(matches("Gene names|normalized Rep")) %>%
  magrittr::set_colnames(c("Gene_names",
                           "A_1000_PFU_PR8_vs_100_PFU_PR8_1",
                           "B_3_×_10^6_CFU_Pseudomonas_vs_1000_PFU_PR8_2",
                           "C_3_×_10^6_CFU_Pseudomonas_vs_100_PFU_PR8_3_invert",
                           "C_3_×_10^6_CFU_Pseudomonas_vs_100_PFU_PR8_1",
                           "A_1000_PFU_PR8_vs_100_PFU_PR8_2_invert",
                           "B_3_×_10^6_CFU_Pseudomonas_vs_1000_PFU_PR8_3_invert",
                           "B_3_×_10^6_CFU_Pseudomonas_vs_1000_PFU_PR8_1",
                           "C_3_×_10^6_CFU_Pseudomonas_vs_100_PFU_PR8_2_invert",
                           "A_1000_PFU_PR8_vs_100_PFU_PR8_3")) %>%
  clean_data() %>% 
  mutate_at(vars(-matches("Gene_names")), list(~log2)) %>% 
  validate_df(valid_list) %>% 
  remove_invalid()

## calculate the p value and group mean
data_201712_stat <-
  map(c("A", "B", "C"),
      function(x) {
        p_value <- sym(glue("{x}_p_value"))
        mean_value <- sym(glue("{x}_mean"))
        grep_string <- glue("^{x}_.+[0-9]$")
        subset <- dplyr::select(data_201712, matches(grep_string))
        
        mutate(
          data_201712,
          !!mean_value := apply(subset, 1, mean),
          !!p_value := apply(subset, 1, function(x)
            t.test(x)$p.value)
        )
      }) %>%
  purrr::reduce(left_join) 

## filter the protein that pass the statistic test
data_201712_stat <-
  map(c("A", "B", "C"),
      function(x) {
        ratio_sig <- sym(glue("{x}_sig"))
        p_value <- sym(glue("{x}_p_value"))
        mean_value <- sym(glue("{x}_mean"))
        
        mutate(data_201712_stat, !!ratio_sig := !!p_value < 0.05 &
                 abs(!!mean_value) > 1)
      }) %>%
  purrr::reduce(left_join) %>% 
  mutate(Sig_diff = rowSums(dplyr::select(., matches("_sig$"))))
```

### Overall impression of datasets

Then we generated a PCA analysis to look at the protein abundance pattern.

```{r PCA-plot-for-the-proteome-involved-bacterial-infection, fig.dim=c(8, 8)}
set.seed(2018)
pca_matrix_201712 <-
  data_201712_stat %>%
  as_tibble() %>%
  dplyr::select(matches("_1|_2|_3")) %>%
  as.matrix %>%
  princomp(., cor = T)

pca_matrix_pov_201712 <- pca_matrix_201712$sdev^2/sum(pca_matrix_201712$sdev^2)

pca_df_201712 <- pca_matrix_201712$loadings[] %>%
  as.data.frame %>%
  mutate(Sample = row.names(.)) %>%
  mutate(Categories = str_extract(Sample, "(?<=[A-Z]{1}_).+(?=_[0-9]{1})") %>% str_replace_all("_", " ")) %>%
  mutate(Batch = str_extract(Sample, "[0-9]{1}$"))

ggplot(pca_df_201712, aes(Comp.1, Comp.2, color = Categories)) +
  geom_point(size = 4) +
  geom_text_repel(aes(label = Batch)) +
  labs(x = sprintf("PC1: %.3f %% Variance", pca_matrix_pov_201712["Comp.1"]*100),
       y = sprintf("PC2: %.3f %% Variance", pca_matrix_pov_201712["Comp.2"]*100)) +
  scale_color_manual(values = brewer.pal(6, "RdBu")[c(1,3, 5)]) +
  theme_classic() +
  theme(aspect.ratio = 1,
        panel.border = element_rect(fill = "transparent",
                                    color = "black"))
```

### Analysis of proteins with significant abundance changes

Lastly, a heatmap for all proteins with different abundance detected in at least one group was visualized.

```{r heatmap-data-preparation}
data_201712_heatmap_ha_df <-
  data.frame(type = colnames(data_201712 %>% dplyr::select(-matches("valid|Gene")))) %>%
  mutate(Comparision = str_extract(type, "(?<=[ABC]_).+(?=_[0-9]{1}$)") %>% 
           str_replace_all("_", " ")) %>%
  dplyr::select(-type)

data_201712_heatmap_ha <-
  columnAnnotation(
    df = data_201712_heatmap_ha_df,
    col = list(
      Comparision = setNames(
        brewer.pal(6, "RdBu")[c(1,3, 5)],
        sort(data_201712_heatmap_ha_df %>% pull(Comparision) %>% unique))),
    annotation_legend_param = list(
      Comparision = list(
        nrow = 3,
        title = "Comparison",
        title_gp = gpar(fontsize = 8),
        labels_gp = gpar(fontsize = 6)
      )
    )
  )
```

The cluster number was first determined by `fviz_nbclust` function with `wss` method.

```{r determine-the-clusters-for-data_201712, fig.dim=c(6, 6)}
data_201712_sig_heatmap_df <-
  data_201712_stat %>%
  dplyr::filter(Sig_diff > 0) %>%
  dplyr::select(matches("Gene|_vs_")) %>%
  `colnames<-`(sapply(colnames(.), function(x) str_replace(x, "^[A-Za-z]{1}_", ""))) %>%
  `colnames<-`(sapply(colnames(.), function(x) str_replace_all(x, "_", " ")))

factoextra::fviz_nbclust(data_201712_sig_heatmap_df %>% dplyr::select(-matches("Gene")),
  kmeans,
  method = "wss"
)
```

Based on the plot, we can see 3 clusters are the best option to visualize the dataset.

```{r heatmap-for-the-proteome-involved-bacterial-infection, fig.dim=c(9, 10)}
set.seed(124)
data_201712_sig_heatmap_df <-
  data_201712_sig_heatmap_df %>%
  mutate(cluster = kmeans(data_201712_sig_heatmap_df %>% dplyr::select(-matches("Gene")), 3)$cluster)

data_201712_sig_stat_heatmap_matrix <-
  data_201712_sig_heatmap_df %>%
  as.matrix() %>%
  `rownames<-`(.[, "Gene names"]) %>%
  .[, !colnames(.) %in% c("Gene names")] %>%
  `class<-`("numeric")

data_201712_sig_stat_heatmap_ha_row <-
  rowAnnotation(
    df = data_201712_sig_heatmap_df %>% dplyr::select(cluster) %>% as.data.frame(),
    col = list(cluster = setNames(
      brewer.pal(3, "BrBG"),
      data_201712_sig_heatmap_df %>% pull(cluster) %>% unique() %>% sort(.)
    )
    ),
    width = unit(0.2, "in"),
    annotation_legend_param = list(
      nrow = 1,
      title = "Cluster",
      title_gp = gpar(fontsize = 8),
      labels_gp = gpar(fontsize = 8),
      legend_width = unit(1, "in")
    )
  )

set.seed(123)
data_201712_sig_stat_heatmap <-
  Heatmap(data_201712_sig_stat_heatmap_matrix[, !colnames(data_201712_sig_stat_heatmap_matrix) %in% c("cluster")],
    top_annotation = data_201712_heatmap_ha,
    heatmap_legend_param = list(
      title = "Log2 Fold Change",
      title_gp = gpar(fontsize = 8),
      labels_gp = gpar(fontsize = 8),
      legend_direction = "horizontal",
      legend_width = unit(1.5, "inch")
    ),
    show_column_names = FALSE,
    show_row_dend = FALSE,
    combined_name_fun = NULL,
    row_names_gp = gpar(fontsize = 8),
    split = data_201712_sig_stat_heatmap_matrix[, "cluster"],
    gap = unit(1.5, "mm")
  )


draw(data_201712_sig_stat_heatmap_ha_row + data_201712_sig_stat_heatmap, heatmap_legend_side = "bottom", annotation_legend_side = "bottom")

for (an in colnames(data_201712_heatmap_ha_df)) {
  decorate_annotation(an, {
    ## annotation names on the left
    grid.text(an,
      unit(1, "npc") + unit(2, "mm"),
      0.6,
      default.units = "npc",
      just = "left",
      gp = gpar(fontsize = 8)
    )
  })
}
```

## State of the machine

```
## sessionInfo()
## R version 3.5.1 (2018-07-02)
## Platform: x86_64-w64-mingw32/x64 (64-bit)
## Running under: Windows 10 x64 (build 17763)
## 
## Matrix products: default
## 
## locale:
## [1] LC_COLLATE=English_Australia.1252  LC_CTYPE=English_Australia.1252   
## [3] LC_MONETARY=English_Australia.1252 LC_NUMERIC=C                      
## [5] LC_TIME=English_Australia.1252    
## 
## attached base packages:
##  [1] parallel  stats4    grid      stats     graphics  grDevices utils    
##  [8] datasets  methods   base     
## 
## other attached packages:
##  [1] blogdown_0.12         ggpubr_0.2            corrplot_0.84        
##  [4] scatterpie_0.1.2      ggthemr_1.1.0         org.Mm.eg.db_3.7.0   
##  [7] AnnotationDbi_1.44.0  IRanges_2.16.0        S4Vectors_0.20.1     
## [10] Biobase_2.42.0        BiocGenerics_0.28.0   customLayout_0.3.0   
## [13] glue_1.3.1            ggalluvial_0.9.1      RColorBrewer_1.1-2   
## [16] egg_0.4.2             cowplot_0.9.4         data.table_1.12.2    
## [19] factoextra_1.0.5      gridExtra_2.3         ComplexHeatmap_1.20.0
## [22] eulerr_5.1.0          ggthemes_4.2.0        ggrepel_0.8.1        
## [25] magrittr_1.5          forcats_0.4.0         stringr_1.4.0        
## [28] dplyr_0.8.0.1         purrr_0.3.2           readr_1.3.1          
## [31] tidyr_0.8.3           tibble_2.1.1          ggplot2_3.1.1        
## [34] tidyverse_1.2.1      
## 
## loaded via a namespace (and not attached):
##  [1] nlme_3.1-137        fs_1.3.1            lubridate_1.7.4    
##  [4] bit64_0.9-7         httr_1.4.0          tools_3.5.1        
##  [7] backports_1.1.4     R6_2.4.0            DBI_1.0.0          
## [10] lazyeval_0.2.2      colorspace_1.4-1    GetoptLong_0.1.7   
## [13] withr_2.1.2         processx_3.3.1      tidyselect_0.2.5   
## [16] bit_1.1-14          compiler_3.5.1      cli_1.1.0          
## [19] rvest_0.3.3         pacman_0.5.1        xml2_1.2.0         
## [22] bookdown_0.10       labeling_0.3        scales_1.0.0       
## [25] callr_3.2.0         digest_0.6.18       rmarkdown_1.12     
## [28] pkgconfig_2.0.2     htmltools_0.3.6     rlang_0.3.4        
## [31] GlobalOptions_0.1.0 readxl_1.3.1        rstudioapi_0.10    
## [34] RSQLite_2.1.1       shape_1.4.4         generics_0.0.2     
## [37] farver_1.1.0        jsonlite_1.6        Rcpp_1.0.1         
## [40] munsell_0.5.0       clipr_0.6.0         whisker_0.3-2      
## [43] stringi_1.4.3       yaml_2.2.0          MASS_7.3-50        
## [46] plyr_1.8.4          blob_1.1.1          crayon_1.3.4       
## [49] lattice_0.20-35     haven_2.1.0         circlize_0.4.6     
## [52] hms_0.4.2           polylabelr_0.1.0    ps_1.3.0           
## [55] knitr_1.22          pillar_1.4.0        rjson_0.2.20       
## [58] reprex_0.2.1        evaluate_0.13       modelr_0.1.4       
## [61] tweenr_1.0.1        cellranger_1.1.0    gtable_0.3.0       
## [64] polyclip_1.10-0     assertthat_0.2.1    xfun_0.6           
## [67] ggforce_0.2.2       broom_0.5.2         rvcheck_0.1.3      
## [70] memoise_1.1.0         
```

